How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-13 doc/paper.tex
grammar checks

We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

@@

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

@@

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. 
%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. 
%Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. 
The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}. The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast forecast the fifth fold.

@@

See our code repository at https://github.com/psmits/trident for full code details. The entire analysis was coded in R and uses tidyverse-style tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}. Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all of the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All of the in-sample AUC estimates from our models are concentrated around an AUC of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

%When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronouced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_full}
%  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
%  \label{fig:auc_taxon}
%\end{figure}

For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

@@

Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.

@@

%When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig. \ref{fig:fold_auc_time}). While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
%  \caption{Comparison of out-of-sample AUC values calculated for each of the My intervals for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds (Fig. \ref{fig:fold_auc}) into the predictions made for each of the million-year intervals. The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
%  \label{fig:fold_auc_time}
%\end{figure}
%We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig. \ref{fig:fold_auc_taxon}). These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups. For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical. In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig. \ref{fig:fold_auc}). Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria. These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g. Foraminifera versus Dinoflagellates.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_full}
%  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models. Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction. Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
%  \label{fig:fold_auc_taxon}
%\end{figure}
%

@@

In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

We quantitatively compared our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). This calculation means that differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and do not form multimillion year patters, quantifying broad congruence that visual comparisons of the two AUC time series suggests (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
  \caption{Comparison between between our out-of-sample forecasts and in-sample forecasts for all models. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figre \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence betwene in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:oos_ins_diff}
\end{figure}

%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
%
%
%% risk estimate compared to change in geo-range
%\begin{figure}[ht]
%  \centering
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
%    \caption{A}
%    \label{fig:relrisk_range}
%  \end{subfigure}
%
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
%    \caption{B}
%    \label{fig:relrisk_ext}
%  \end{subfigure}
%  \caption{Comparison between species geogrpahic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
%  \label{fig:relrisk_compare}
%\end{figure}

@@

One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

@@

The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.

@@

Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

@@

Temperature was also transformed and standardized the in the same manner as geographic range. Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.

@@

Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates. The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model. The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\). The difference in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations. The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla. The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}. This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla. \(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum. \(A\) is modeled as following a multivariate normal distribution with phylum means being vector \(\delta\) and covariance matrix \(\Sigma_{A}\). The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).

>>>>

We cannot know, however, we do not yet know which species are going to go extinct because this has not happened yet -- it is unobservable. We approach this problem by analyzing the past in order to predict the future. The fossil record preserves past extinction events, allowing us to develop a predictive model of species extinction based on this record and the properties of the observed species, both extinct and extant \citep{Harnik2012a,Finnegan2015}. By assessing the predictive performance of this model on unobserved data, we can quantify how precise our best estimates will be for future extinctions -- we ask the probability that, given two random species, we correctly rank their relative risks of extinction.

@@

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

@@

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. 
%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. 
%Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. 
The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}. The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast the fifth fold.

@@

See our code repository at https://github.com/psmits/trident for full code details. The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}. Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

%When the posterior predictive distributions of the in-sample AUC estimates are presented by taxonomic group, some heterogeneity in model adequacy is revealed (Fig. \ref{fig:auc_taxon}). While in all cases the model with the highest average in-sample AUC is the parameter-rich ``past and vary'' model, the amount of difference between the models varies by taxonomic group in ways not observable from the pooled estimates (Fig. \ref{fig:auc_hist}). For example, the difference between the ``past and vary'' model and the others is more pronounced for Calcareous nannoplankton and Dinoflagellates, and smaller for the Foraminifera and Radiolaria. 
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_full}
%  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups present in this analysis. The densities reflect the posterior distribution of the estimates, and below each density is marked the median AUC value along with the 50\% and 80\% credible intervals. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values.}
%  \label{fig:auc_taxon}
%\end{figure}

There are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 for the various taxon/model combinations (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

@@

Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.

@@

%When the posterior predictive distribution of expected out-of-sample AUC is presented as a time series, the similarity between the models is even more apparent (Fig. \ref{fig:fold_auc_time}). While the width of the credible intervals at various time points varies between the models, the overall picture of expected out-of-sample AUC is almost identical when you compare the models -- periods of relatively better or worse performance map identically between the time series.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time_full}
%  \caption{Comparison of out-of-sample AUC values calculated for each of the million year intervals for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds (Fig. \ref{fig:fold_auc}) into the predictions made for each of the million-year intervals. The red line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend.}
%  \label{fig:fold_auc_time}
%\end{figure}
%We can also compare expected out-of-sample AUC by taxonomic group for each of the models (Fig. \ref{fig:fold_auc_taxon}). These comparisons reveal a lot about the differences in predictive potential of the taxonomic groups. For example, the posterior predictive distributions of out-of-sample AUC for Foraminifera from all four models are approximately identical. In contrast, expected out-of-sample AUC for Radiolaria exhibits the same or similar pattern in relative model performance to the pooled comparisons (Fig. \ref{fig:fold_auc}). Additionally, we can state that our out-of-sample predictions for calcareous nannoplankton and dinoflagellates are not necessarily as precise as our estimates for Foraminifera or even Radiolaria. These results indicate that out-of-sample predictions may be easier for some taxonomic groups than others (e.g. Foraminifera versus Dinoflagellates.
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_full}
%  \caption{Comparison of out-of-sample AUC values aggregated by taxonomic group for each of the four models. Depicted for each taxon-model combination is an aggregate density of all posterior predictive estimates for each of four folds -- cross-validation estimates are commonly multi-modal as each fold presents its own challenges for prediction. Beneath these densities is marked the median estimate along with 50\% and 80\% credible intervals.}
%  \label{fig:fold_auc_taxon}
%\end{figure}
%

@@

In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

We quantitatively compared our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). This calculation means that differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, quantifying the broad congruence that the visual comparison of the two types of AUC time series suggest (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_diff}
  \caption{Comparison between our out-of-sample forecasts and in-sample forecasts for all models. This value is calculated as the values presented in Figure \ref{fig:fold_auc_taxon_time} minus those values presented in Figure \ref{fig:auc_taxon_time}. A differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts, while a positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, and a negative difference indicates poorer out-of-sample performance than in-sample forecast. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:oos_ins_diff}
\end{figure}

%Looking at four randomly sampled species, we evaluated their probability of extinction at all times that species was observed. We then compared these estimates to the geographic range trajectory of 
%
%
%% risk estimate compared to change in geo-range
%\begin{figure}[ht]
%  \centering
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_range}
%    \caption{A}
%    \label{fig:relrisk_range}
%  \end{subfigure}
%
%  \begin{subfigure}{\textwidth}
%    \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/relrisk_ext}
%    \caption{B}
%    \label{fig:relrisk_ext}
%  \end{subfigure}
%  \caption{Comparison between species geographic range histories and our estimate of their probability of extinction for each time of observation. Geographic range is assumed observed without error. Our probability estimates are presented as a median (point) and 50\% and 80\% credible intervals.}
%  \label{fig:relrisk_compare}
%\end{figure}

@@

One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

@@

The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.

@@

Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

@@

Temperature was also transformed and standardized the in the same manner as geographic range. Difference in temperature between observations was measured from the standardized temperature values and was not standardized separately.

@@

Our covariates matrix \(X\) is a \(N \times D\) matrix where \(N\) is the total number of observations and \(D\) is the total number of covariates. The first column of \(X\) is entirely 1's as it corresponds to the intercept term in the regression model. The next two columns of \(X\) are two aspects of geographic range as continuous covariates: geographic range \(r\) during interval \(t\), and the difference \(d\) between the geographic range at \(t - 1\) and \(t\). Change in geographic range was calculated from the transformed and standardized geographic range values; this means that change in geographic range is in units of changes in standard deviations. The final two columns are two aspects of global temperature: mean temperature during interval \(t\), and the lag of mean temperature (i.e. mean temperature during interval \(t - 1\).) As with change to geographic range, the lag of temperature is based on the transformed and standardized temperature estimates.

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla. The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each Coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}. This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla. \(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum. \(A\) is modeled as following a multivariate normal distribution with phylum means being the vector \(\delta\) and covariance matrix \(\Sigma_{A}\). The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).
